Efficient and Scalable Bipartite Matching with Fast Beta Linkage (fabl)

### 2023-07-19 ba_resubmit/response_to_reviewers2.tex

	\pointRaised{R}{%
	For the missing data treatment now included just before Section 2.1, is the assumption truly missing at random, or missing completely at random? Or, does this distinction not matter because of the independence assumed across elements of the comparison vector?}

	%% RCS: See two comments below regarding MAR versus MCAR. I have an alternative response for the one regarding miss at random.

%	\textcolor{teal}{Answer for MAR: Yes, the assumption is truly missing at random. This distinction between truly missing at random and missing completely at random does not matter due to two assumptions. Following Sadinle (2017), we assume that the fields in the comparison vectors are conditionally independent given the coreference matrix and we assume that the missing comparisons are ignorable (Sadinle (2017). Therefore, our likelihood is the same as in Sadinle (2017) (see equations (3) and (4) of his paper). As stated by Sadinle (2017), ``we work under the assumption of ignorability of the missingness mechanism for the comparisons so that we can base our inferences on the marginal distribution of the observed comparisons (Little and Rubin, p. 90)."  We refer to Sadinle (2017), Section 4.2, page 604 for further details.}

	\reply{The assumption is missing completely at random (MCAR) due to fact that we assume the fields in the comparison vectors are conditionally independent given the coreference matrix (or matching label). The assumption is not missing at random (MAR) as we do not assume that the missing comparisons are also ignorable. See Section 3, page 5 regarding clarification of this point \citep{LittleRubin2002}.}
	\textcolor{red}{Still checking on this and looking at Little and Rubin.}

### 2023-07-19 ba_resubmit/kundinger_resubmit2.tex

Many probabilistic record linkage methods are rely on the seminal work of \cite{fellegi_theory_1969} and \cite{newcombe_automatic_1959}. In their approach, the data analyst first creates comparison vectors for each pair of records in the data files. These vectors indicate how similar the records are on a set of variables measured in both files, known as the linkage variables.  Using these comparison vectors, the analyst classifies each pair as a match or nonmatch using a likelihood ratio test. Crucially, these decisions are made independently for each pair. The \cite{fellegi_theory_1969} approach has been extended for a wide variety of applications \citep[e.g.,][]{Winkler1990, fair2004generalized, wagner2014person, gill2003english, enamorado2019using, aleshinguendel2021multifile}. An alternative paradigm is to model the linkage variables directly \citep[e.g.,][]{tancredi2011hierarchical, steorts_bayesian_2016, marchant_distributed_2019, betancourt2021prior}. In this article, we build on the contributions to the comparison vector approach.

@@

In this article, we propose fast beta linkage (\texttt{fabl}), which extends the \texttt{BRL} model for increased efficiency and scalability. {\color{red}{We use independent priors for the matching status of each record, inducing a dependency that is stronger than the record pair independence in the original \cite{fellegi_theory_1969} but weaker that the one-to-one requirement of \cite{sadinle_bayesian_2017}.}} \textcolor{teal}{This independence assumption allows for a scalable update of the Gibbs sampler, contrasting sequential updates of \cite{sadinle_bayesian_2017}. We employ the decision theoretic technique of \cite{sadinle_bayesian_2017} and \cite{steorts_bayesian_2016} in order to ensure our method after record linkage is bipartite.}
This \textcolor{teal}{discovery} allows us to (1) employ hashing techniques that speed up calculations and reduce computational costs, (2) compute the pairwise record comparisons over large data files via parallel computing, and (3) reduce memory costs through what we call storage efficient indexing. These contributions allow \texttt{fabl} to perform record linkage on much larger data files than previous Bayesian Fellegi-Sunter models at significantly increased speed with similar levels of accuracy. In particular, computation time under \texttt{BRL} grows quadratically, with the size of each data file, while computation time under \texttt{fabl} grows linearly, only with the size of the smaller data file.

@@

Instead of the prior over $\bm{Z}$ from \cite{sadinle_bayesian_2017}, we follow \cite{wortman2019} and use independent priors for each component $Z_j$. However, unlike \cite{wortman2019} who proposes a flat prior for $Z_j$, 
\textcolor{teal}{we use propose an alternative, called the fast beta linkage prior (fabl).}
%the fast Beta prior as follows. 
For each $Z_j$, we have
\begin{subequations}
\begin{align}
	p(Z_j = q| \pi) &= \begin{cases} 
	\frac{1}{n_A}\pi,  & q \leq n_A;\\
	1-\pi, &  q  = n_A + j;  \\
\end{cases} \label{eqn:fast_beta_prior}  \\
\pi &\sim \text{Beta}(\alpha_{\pi}, \beta_{\pi}) \label{eqn:fast_beta_prior2}.
\end{align}
\end{subequations}
We can interpret (\ref{eqn:fast_beta_prior}) as follows: record $B_j$ has some match in $A$ with probability $\pi$, and each record $A_i$ is equally likely to be that match. The hyperparameters $\alpha_{\pi}$ and $\beta_{\pi}$ encode prior beliefs about  the proportion of records in $B$ that have matches in $A.$

@@

\begin{subequations}
\begin{align}
	\mathcal{L}(\bm{Z}, \bm{m}, \bm{u} \mid \gamma) &= \prod_{i=1}^{n_A}  \prod_{j=1}^{n_B}\prod_{f=1}^{F}\prod_{l=1}^{L_f}\left[  m_{fl}^{I(Z_j = i)}u_{fl}^{I(Z_j \neq i)}\right]^{I(\gamma_{ij}^f = l)I_{obs}(\gamma_{ij}^f)}, \label{eqn:likelihood}\\
	\bm{m}_f &\sim \text{Dirichlet}(\alpha_{f1}, \ldots, \alpha_{f L_f}), \forall f = 1, \ldots, F, \label{eqn:m} \\
	\bm{u}_f &\sim \text{Dirichlet}(\beta_{f1}, \ldots, \beta_{f L_f}),\forall f = 1, \ldots, F,  \label{eqn:u}\\
	p(Z_j = q| \pi)  &=
	\begin{cases} 
		\frac{1}{n_A}\pi,  & q \leq n_A; \\
		1-\pi, &  q  = n_A + j; \\
	\end{cases} \label{eqn:z}\\
	\pi &\sim \text{Beta}(\alpha_{\pi}, \beta_{\pi})\label{eqn:pi},
\end{align}
\end{subequations}
\textcolor{teal}{where $\alpha_{\pi}, \beta_{\pi}$ are fixed hyper-parameters.}

@@

We \textcolor{teal}{approximate} the posterior distribution of the parameters in (\ref{eqn:likelihood} - \ref{eqn:pi}) using a Gibbs sampler. 
We initialize $\bm{m}$ and $\bm{u}$ from random draws from their prior distributions, and initialize $\bm{Z}$ to reflect no matches across data files; that is, $\bm{Z} = (n_A + 1, \ldots, n_A + n_B)$. To take the $(s+1)$th sample of each $\bm{m}_f$ and $\bm{u}_f$ given $\bm{Z}^{(s)}$, we use the full conditionals,

\textcolor{teal}{We present a Gibbs sampler to approximate the joint posterior distribution of $\bm{Z}, \bm{m}, \bm{u}$ given the observed comparison data from the likelihood and priors assumed in (\ref{eqn:likelihood} - \ref{eqn:pi}). While it is easy to marginalize over $\pi$ and derive a collapsed Gibbs sampler, the sampler is sequential in nature as is that of \cite{}. Thus, we proceed with a traditional Gibbs sampler, where we show in Appendix \ref{}, the updates are parallel, making our approach scalable.}

@@

\textcolor{teal}{This implies that we can make parallel updates to $\bm{Z}$ instead of the traditional sequential updates as in \cite{sadinle_bayesian_2017}.
In fact, to our knowledge, our approach and that of \cite{wortman2019} are the only two Bayesian approaches that propose parallel updates to the matching label (or linkage structure), contrasting the Bayesian record linkage literature \citep{tancredi2011hierarchical, sadinle_detecting_2014, steorts_bayesian_2016, marchant_distributed_2019, zanella_flexible_2016, betancourt2022prior, aleshinguendel2021multifile}. See Appendix~\ref{app:derivations} for derivations of the full conditional distributions.}

@@

\textcolor{teal}{Finally, to obtain an estimate $\hat{\bm{Z}}$ of the linkage structure, we use the loss functions and Bayes estimate from \cite{sadinle_bayesian_2017}. However, since (\ref{eqn:fast_beta_prior}) does not strictly enforce one-to-one matching, it is possible for this Bayes estimate to link multiple records in $B$ to one record in $A$. To obtain a Bayes estimate that fulfills the one-to-one matching (or bipartite) requirement, we minimize the expected loss subject to the constraint that $\hat{Z}_j \neq \hat{Z}_{j'}$ for all $j \neq j'$. Details for the initial Bayes estimate and the post-processing procedure are provided in Appendix \ref{bayes-estimate}.}

>>>>

Many probabilistic record linkage methods are rely on the seminal work of \cite{fellegi_theory_1969} and \cite{newcombe_automatic_1959}. In their approach, the data analyst first creates comparison vectors for each pair of records in the data files. These vectors indicate how similar the records are on a set of variables measured in both files, known as the linkage variables.  Using these comparison vectors, the analyst classifies each pair as a match or nonmatch using a likelihood ratio test. Crucially, these decisions are made independently for each pair. The \cite{fellegi_theory_1969} approach has been extended for a wide variety of applications \citep[e.g.,][]{Winkler1990, fair2004generalized, wagner2014person, gill2003english, enamorado2019using, aleshinguendel2021multifile}. An alternative paradigm is to model the linkage variables directly \citep[e.g.,][]{tancredi2011hierarchical, steorts_bayesian_2016, marchant_distributed_2019, betancourt2022prior}. In this article, we build on the contributions to the comparison vector approach.

@@

In this article, we propose fast beta linkage (\texttt{fabl}), which extends the \texttt{BRL} model for increased efficiency and scalability. {\color{red}{We use independent priors for the matching status of each record, inducing a dependency that is stronger than the record pair independence in the original \cite{fellegi_theory_1969} but weaker that the one-to-one requirement of \cite{sadinle_bayesian_2017}.}} \textcolor{teal}{This independence assumption allows for a scalable update of the Gibbs sampler, contrasting sequential updates of \cite{sadinle_bayesian_2017}. We employ the decision theoretic technique of \cite{sadinle_bayesian_2017} in order to ensure our method after record linkage is bipartite.}
This \textcolor{teal}{discovery} allows us to (1) employ hashing techniques that speed up calculations and reduce computational costs, (2) compute the pairwise record comparisons over large data files via parallel computing, and (3) reduce memory costs through what we call storage efficient indexing. These contributions allow \texttt{fabl} to perform record linkage on much larger data files than previous Bayesian Fellegi-Sunter models at significantly increased speed with similar levels of accuracy. In particular, computation time under \texttt{BRL} grows quadratically, with the size of each data file, while computation time under \texttt{fabl} grows linearly, only with the size of the smaller data file.

@@

Instead of the prior over $\bm{Z}$ from \cite{sadinle_bayesian_2017}, we follow \cite{wortman2019} and use independent priors for each component $Z_j$. However, unlike \cite{wortman2019} who proposes a flat prior for $Z_j$, 
\textcolor{teal}{we propose an alternative, called the fast beta linkage (fabl) prior.}
%the fast Beta prior as follows. 
For each $Z_j$, we have
\begin{subequations}
\begin{align}
	p(Z_j = q| \pi) &= \begin{cases} 
	\frac{1}{n_A}\pi,  & q \leq n_A;\\
	1-\pi, &  q  = n_A + j;  \\
\end{cases} \label{eqn:fast_beta_prior}  \\
\pi &\sim \text{Beta}(\alpha_{\pi}, \beta_{\pi}) \label{eqn:fast_beta_prior2}.
\end{align}
\end{subequations}
We can interpret (\ref{eqn:fast_beta_prior}) as follows: record $B_j$ has some match in $A$ with probability $\pi$, and each record $A_i$ is equally likely to be that match. The hyperparameters $\alpha_{\pi}$ and $\beta_{\pi}$ encode prior beliefs about  the proportion of records in $B$ that have matches in $A.$

@@

\begin{subequations}
\begin{align}
	\mathcal{L}(\bm{Z}, \bm{m}, \bm{u} \mid \gamma) &= \prod_{i=1}^{n_A}  \prod_{j=1}^{n_B}\prod_{f=1}^{F}\prod_{l=1}^{L_f}\left[  m_{fl}^{I(Z_j = i)}u_{fl}^{I(Z_j \neq i)}\right]^{I(\gamma_{ij}^f = l)I_{obs}(\gamma_{ij}^f)}, \label{eqn:likelihood}\\
	\bm{m}_f &\sim \text{Dirichlet}(\alpha_{f1}, \ldots, \alpha_{f L_f}),\quad  \forall  \; f = 1, \ldots, F, \label{eqn:m} \\
	\bm{u}_f &\sim \text{Dirichlet}(\beta_{f1}, \ldots, \beta_{f L_f}), \quad \forall \; f = 1, \ldots, F,  \label{eqn:u}\\
	p(Z_j = q| \pi)  &=
	\begin{cases} 
		\frac{1}{n_A}\pi,  & q \leq n_A; \\
		1-\pi, &  q  = n_A + j; \\
	\end{cases} \label{eqn:z}\\
	\pi &\sim \text{Beta}(\alpha_{\pi}, \beta_{\pi})\label{eqn:pi},
\end{align}
\end{subequations}
\textcolor{teal}{where recall that $\alpha_{\pi}, \beta_{\pi}$ are fixed hyper-parameters.}

@@

\textcolor{teal}{We present a Gibbs sampler to approximate the joint posterior distribution of $\bm{Z}, \bm{m},$ and  $\bm{u}$ given the observed comparison data from the likelihood and priors assumed in (\ref{eqn:likelihood} - \ref{eqn:pi}). While we could marginalize over $\pi$ and derive a collapsed Gibbs sampler, in the spirit of \cite{sadinle_bayesian_2017}, this leads to sequential updates, which are computationally slow. Instead, we proceed with a traditional Gibbs sampler, where the updates to the matching label (or coreference matrix) are parallel, allowing for more efficient updates. Refer to Appendix~\ref{app:derivations} for derivations of the full conditional distributions.}

@@

\textcolor{teal}{This implies that we can make parallel updates to $\bm{Z}$ instead of the traditional sequential updates as in \cite{sadinle_bayesian_2017}.
In fact, to our knowledge, our approach and that of \cite{wortman2019} are the only two Bayesian approaches that propose parallel updates to the matching label (or linkage structure), contrasting the Bayesian record linkage literature \citep{tancredi2011hierarchical, sadinle_detecting_2014, steorts_bayesian_2016, marchant_distributed_2019, zanella_flexible_2016, betancourt2022prior, aleshinguendel2021multifile}. }
%See Appendix~\ref{app:derivations} for derivations of the full conditional distributions.}

@@

\textcolor{teal}{Finally, to obtain an estimate $\hat{\bm{Z}}$ of the linkage structure, we use the loss functions and Bayes estimate from \cite{sadinle_bayesian_2017}. However, since (\ref{eqn:fast_beta_prior}) does not strictly enforce one-to-one matching, it is possible for this Bayes estimate to link multiple records in $B$ to one record in $A$. To obtain a Bayes estimate that fulfills the one-to-one matching (or bipartite) requirement, we minimize the expected loss subject to the constraint that $\hat{Z}_j \neq \hat{Z}_{j'}$ for all $j \neq j'$. See Appendix \ref{bayes-estimate} for details regarding the initial Bayes estimate and the post-processing procedure.}

%Details for the initial Bayes estimate and the post-processing procedure are provided in Appendix \ref{bayes-estimate}.
